Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2018-10-02 paper/paper.tex
some very minor tweaks to Intro text

It remains unclear what determines if an infected cell triggers an innate-immune response.
Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
For instance, only some cells induce IFN even upon homogeneous treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and there is evidence that the frequency of IFN induction depends on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
But for influenza, a third possible contributor also looms large: viral genetic diversity.
Specifically, because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wildtype,'' individual virions will often have genetic defects~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\citep{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and more IFN is induced by viral stocks that contain defective virions with internal deletions in the polymerase genes~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.

However, existing single-cell techniques are inadequate to determine how viral genetic diversity actually contributes to cell-to-cell heterogeneity during infection, since they do not systematically determine the full sequences of the virions that infect individual cells.
Flow cytometry and fluorescent reporters only measure expression of viral proteins~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of viral RNAs~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous}.
None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations will be common regardless of whether the cells are infected with a nominally ``wildtype'' or ``mutant'' stock of virus.

Here we develop a new approach to measure both the full transcriptome and the sequences of all viral genes in single influenza-infected cells that do and do not activate innate immunity.
To do this, we first enrich for rare IFN+ cells, and then perform both standard Illumina-based single-cell transcriptomics and full-length PacBio sequencing of viral genes.
We obtain the transcriptomes and full sequences of all expressed viral genes in 150 virus-infected cells, 49 of which express IFN.
Even though we used a nominally ``wildtype'' stock of virus, two-thirds of the cells are infected by virions that have a mutation or defect in gene expression.
This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by truly wildtype virions having a tighter distribution of viral transcriptional burden.
Viral defects are especially common in IFN+ cells, and we identify at least four defects that are causally associated with IFN induction: amino-acid substitutions in the viral polymerase (PB1), absence of the NS gene, amino-acid substitutions in NS1, and internal deletions in polymerase genes.
However, viral genetic variation does not fully explain the cell-to-cell heterogeneity, as even truly wildtype virions sometimes induce IFN, and no viral mutations that we identify deterministically induce IFN.

>>>>

It remains unclear why only some infected cells trigger innate-immune responses.
Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
For instance, only some cells induce IFN even upon homogeneous treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and there is evidence that the frequency of IFN induction depends on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
But for influenza, a third possible contributor also looms large: viral genetic diversity.
Specifically, because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wildtype,'' individual virions will often have genetic defects~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\citep{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and more IFN is induced by viral stocks that contain defective virions with internal deletions in the polymerase genes~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.

However, existing single-cell techniques are inadequate to determine how viral genetic diversity actually contributes to cell-to-cell heterogeneity during infection, since they do not systematically determine the full sequences of the virions that infect individual cells.
Flow cytometry and fluorescent reporters only measure expression of viral proteins~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of viral transcripts~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous}.
None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations will be common regardless of whether the cells are infected with a nominally ``wildtype'' or ``mutant'' stock of virus.

Here we develop a new approach to measure both the full transcriptome and the sequences of all viral genes in single influenza-infected cells.
To do this, we first enrich for rare IFN+ cells, and then perform both standard Illumina-based single-cell transcriptomics and full-length PacBio sequencing of viral genes.
We obtain the transcriptomes and full sequences of all expressed viral genes in 150 virus-infected cells, 49 of which express IFN.
Even though we used a nominally ``wildtype'' stock of virus, two-thirds of the cells are infected by virions that have a mutation or defect in gene expression.
This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by truly wildtype virions having a tighter distribution of viral transcriptional burden.
Viral defects are especially common in IFN+ cells, and we identify at least four defects that cause increased IFN induction: amino-acid substitutions in the viral polymerase (PB1), absence of the NS gene, amino-acid substitutions in NS1, and internal deletions in polymerase genes.
However, viral genetic variation does not fully explain the cell-to-cell heterogeneity, as even truly wildtype virions sometimes induce IFN, and no viral mutations that we identify deterministically induce IFN.
Therefore, viral diversity is an important but not exclusive cause of cell-to-cell heterogeneity during influenza infection.
